A coherent and comprehensive dating system for stars in the Milky Way

### 2017-05-09 paper/ms.tex
editing method section and adding more words

We combine the following dating methods: gyrochronology, isochrone fitting,
asteroseismology and galactic kinematics into one model.
In classical isochrone fitting, one might calculate the probability of age
($A$), mass ($M$), \feh, distance ($D$) and extinction (\av), given a set of
apparent magnitudes in different filters, \eg\ $M_V$, $M_J$ and $M_K$
\begin{equation}
    p(A, M, [Fe/H], D, A_v | M_V, M_J, M_K) \propto p(M_v, M_J, M_K |
    A, M, [Fe/H], D, A_v)p(A, M, [Fe/H], D, A_v).
\end{equation}
Note that there are no free parameters in the model here --- the model is
entirely governed by physics and does not need to be calibrated further.
Similarly, in classical gyrochronology, one might calculate the probability of
age given rotation period and $B-V$ colour,
\begin{equation}
    p(A | P_{rot}, B-V) \propto p(P_{rot}, B-V | A)p(A)
\end{equation}
In a gyrochronology model, because our understanding of the physics is
incomplete, there are some free parameters that must be calibrated.
In the \citet{Barnes2003} gyrochronology parameterisation, these are $a$, $b$
and $n$, $A = P^{1/n} - (B-V-0.4)^{b/n}$.
During a calibration exercise, the observables, $A$, $P_{rot}$ and $B-V$ will
be known and the parameters, $a$, $b$ and $n$ will be inferred,
\begin{equation}
    p(a, b, n | A, P_{rot}, B-V) \propto p(A, P_{rot}, B-V | a,
    b, n) p(a, b, n).
\end{equation}
The vertical action-age relation also has free parameters that must be
calibrated using observations, $A = \alpha (\sigma_{Jz}^2)^\beta$.
These free parameters, $\alpha$ and $\beta$ can, again be calibrated via
\begin{equation}
    p(\alpha, \beta | \{A\}, \{J_z\}) \propto p(\{A\}, \{J_z\} | \alpha,
    \beta) p(\alpha, \beta).
\end{equation}

The unique aspect of our analysis is to combine inference over ages with
calibration of the free parameters in the gyrochronology and vertical
action-age relations:
\begin{eqnarray}
    &&p(\{A\}, \{M\}, \{[Fe/H]\}, \{D\}, \{A_v\}, a, b, n, \alpha, \beta|
    \{M_V\}, \{M_J\}, \{M_K\}, \{P_{rot}\}, \{B-V\}, \{J_z\}) \\
    \propto &&p(\{M_v\}, \{M_J\}, \{M_K\}, \{P_{rot}\},
    \{B-V\}, \{J_z\} | \{A\}, \{M\}, \{[Fe/H]\}, \{D\}, \{A_v\}, a, b, n,
    \alpha, \beta) \\
    &&p(A, M, [Fe/H], D, A_v, a, b, n, \alpha, \beta).
\end{eqnarray}
We can use conditional independence to break this down further into the
following:
\begin{eqnarray}
    &&p(\{A\}, \{M\}, \{[Fe/H]\}, \{D\}, \{A_v\}, a, b, n, \alpha, \beta|
    M_V, M_J, M_K,
    \{P_{rot}\}, \{B-V\}, \{J_z\}) \\
    \propto &&p(\{M_v\}, \{M_J\}, \{M_K\} | \{A\}, \{M\}, \{[Fe/H]\}, \{D\},
    \{A_v\})
    p(A)p(M)p([Fe/H])p(D)p(A_v) \\
    &&p(\{P_{rot}\}, \{B-V\}, a, b, n | A)p(a)p(b)p(n) \\
    &&p(\{J_z\}, \alpha, \beta | \{A\}) p(\alpha)p(\beta).
\end{eqnarray}
In other words
\begin{eqnarray}
    &&p(\mathrm{Ages~\&~other~parameters}|\mathrm{observables}) \\
    \propto &&\mathcal{L}_{\mathrm{isochronal~age}} \times
    \mathcal{L}_{\mathrm{gyro~age}} \times
    \mathcal{L}_{\mathrm{kinematic~age}} \\
    && \times p(\mathrm{Ages~\&~other~parameters})
\end{eqnarray}.

>>>>

We combine the following dating methods: gyrochronology, isochrone fitting,
asteroseismology and galactic kinematics into one model.
In classical isochrone fitting, one might calculate the probability of age
$A$, mass $M$, \feh, distance $D$ and extinction \av, given a set of
apparent magnitudes in different filters, \eg\ $M_V$, $M_J$ and $M_K$.
This can be written
\begin{eqnarray}
    &&p(A, M, [Fe/H], D, A_v | M_V, M_J, M_K) \\ \nonumber
    \propto &&p(M_v, M_J, M_K | A, M, [Fe/H], D, A_v) \\ \nonumber
    &&p(A)p(M)p([Fe/H])p(D)p(A_v).
\end{eqnarray}
There are no free parameters in the model here --- the model is entirely
determined by physics and does not need to be calibrated further.
Similarly, in classical gyrochronology, one might calculate the probability of
age given rotation period and $B-V$ colour,
\begin{equation}
    p(A | P_{rot}, B-V) \propto p(P_{rot}, B-V | A)p(A)
\end{equation}
In a gyrochronology model however our understanding of the physics is
incomplete and some free parameters must be calibrated.
In the \citet{Barnes2003} gyrochronology parameterisation, $A =
P_{\mathrm{rot}}^{1/n} - (B-V-0.4)^{b/n}$, these free parameters are $a$, $b$
and $n$.
During a calibration exercise, the observables, $A$, $P_{rot}$ and $B-V$ will
be known and the parameters, $a$, $b$ and $n$ will be inferred,
\begin{equation}
    p(a, b, n | \{A\}, \{P_{rot}\}, \{B-V\}) \propto \{p(A\}, \{P_{rot}\},
    \{B-V\} | a, b, n) p(a)p(b)p(n).
\end{equation}
The vertical action-age relation, $A = \alpha (\sigma_{Jz}^2)^\beta$,
also has free parameters that must be calibrated using observations.
These free parameters, $\alpha$ and $\beta$ can again be calibrated via
\begin{equation}
    p(\alpha, \beta | \{A\}, \{J_z\}) \propto p(\{A\}, \{J_z\} | \alpha,
    \beta) p(\alpha, \beta).
\end{equation}

The unique aspect of our analysis is to combine inference over ages with
calibration of the free parameters in the gyrochronology and vertical
action-age relations:
\begin{eqnarray}
    &&p(\{A\}, \{M\}, \{[Fe/H]\}, \{D\}, \{A_v\}, a, b, n, \alpha, \beta|
    \{M_V\}, \{M_J\}, \{M_K\}, \{P_{rot}\}, \{B-V\}, \{J_z\}) \\ \nonumber
    \propto &&p(\{M_v\}, \{M_J\}, \{M_K\}, \{P_{rot}\},
    \{B-V\}, \{J_z\} | \{A\}, \{M\}, \{[Fe/H]\}, \{D\}, \{A_v\}, a, b, n,
    \alpha, \beta) \\ \nonumber
    &&p(A, M, [Fe/H], D, A_v, a, b, n, \alpha, \beta).
\end{eqnarray}
We can use conditional independence to break this down further into the
following:
\begin{eqnarray}
    &&p(\{A\}, \{M\}, \{[Fe/H]\}, \{D\}, \{A_v\}, a, b, n, \alpha, \beta|
    M_V, M_J, M_K,
    \{P_{rot}\}, \{B-V\}, \{J_z\}) \\  \nonumber
    \propto &&p(\{M_v\}, \{M_J\}, \{M_K\} | \{A\}, \{M\}, \{[Fe/H]\}, \{D\},
    \{A_v\})
    p(A)p(M)p([Fe/H])p(D)p(A_v) \\ \nonumber
    &&p(\{P_{rot}\}, \{B-V\}, a, b, n | A)p(a)p(b)p(n) \\ \nonumber
    &&p(\{J_z\}, \alpha, \beta | \{A\}) p(\alpha)p(\beta).
\end{eqnarray}
In other words
\begin{eqnarray}
    &&p(\mathrm{Ages~\&~other~parameters}|\mathrm{observables}) \\
    \propto &&\mathcal{L}_{\mathrm{isochronal~age}} \times
    \mathcal{L}_{\mathrm{gyro~age}} \times
    \mathcal{L}_{\mathrm{kinematic~age}} \\
    && \times p(\mathrm{Ages~\&~other~parameters})
\end{eqnarray}.
We can multiply the likelihoods together by making the assumption that the
probabilities are conditionally independent, \ie\ a gyrochronology age depends
only on rotation period and colour, not proper motion, for example.
In reality this assumption is not valid because gyrochronology ages do depend
on \logg, \teff\ and \feh.
\racomment{How do we get around this!?}

We explore the parameter space via Gibbs sampling.
Gibbs sampling is a variation of MCMC that makes use of conditional
independence within the problem to reduce dimensionality.
It is often used in hierarchical inference problems such as this, where one
has a set of global parameters ($a$, $b$, $n$, $\alpha$ \& $\beta$) and sets
of per-object parameters ($\{A\}$, $\{M\}$, $\{[Fe/H]\}$, $\{D\}$, $\{A_v\}$).
Instead of updating all parameters simultaneously as in Metropolis,
Metropolis-Hastings, affine invariant \citep{Goodman2010}, nested, Hamiltonian,
and many other sampling algorithms, in Gibbs sampling, subsets of parameters
are updated.
For example, the gyrochronology parameters will be updated, then the kinematic
parameters, then the isochronal parameters for star 1, then star 2, etc.
Since the number of parameters is $N_{gyro} + N_{kinematic} + N_{iso} \times
N_{stars}$, the number of parameters grows rapidly as stars are added.
